Notes

### 2022-09-21 notes/notes.tex
update notation

Now we present a different interpretation. Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x}_{365} := S^{-1} x \otimes \vb{1}_{365}$. This is a vector of length $365 \times 200$ that gives the emission from day 1 to day 365 for each source. This is assumed to be constant for all days. Therefore, $\tilde{x}_{365}$ consists of $m = 200$ blocks, each of which is a constant vector of length $365$.

We can write a system of equations that is logically the same as \cref{eq: initial model concise} but uses $\tilde{x}_{365}$. This gives a larger linear system. Because the emission from most dates has a negligible contribution to the observed concentrations, many coefficients in the larger matrix will be zero. For every measurement $i$, we have a linear relationship that looks something like this:
$$
        \hat{y}_i
    =
    \begin{bmatrix}
        \cdots & M_{i,(1,1)} & \dots & M_{i,(1,15)} & 0 & \dots & M_{i,(2,1)} & \dots & M_{i,(2,15)} & \dots
    \end{bmatrix}
    (S^{-1} x \otimes \vb{1}_{365})
.
$$
The presence of zeros in the matrix indicates that emissions from the distant past do not contribute to the observed concentrations.

Now assume that in each station, there is one observation per day, named $y_{kt}$, where $k = 1,\dots,K$ where $K$ is the number of stations and $t = 1,\dots,365$. The index $t=1$ corresponds to the most recent measurement. Then the full system reads as follows:
\begin{equation}
    \label{eq: initial model 365 days}
\begin{bmatrix}
    \hat{y}_{1,1} \\
    \hat{y}_{1,2} \\
    \vdots \\
    \hat{y}_{2,1} \\ 
    \vdots \\
    \hat{y}_{K,365}
\end{bmatrix}
=
\underbrace{
\begin{bmatrix}
    M_{(1,1), (1,1)} & M_{(1,1), (1,2)} & \cdots & M_{(1,1),(1,15)} & 0 & \cdots & \cdots & M_{(1,1), (2,1)} & \cdots & 0 \\
    0 & M_{(1,2), (1,1)} & \cdots & M_{(1,2), (1,14)} & M_{(1,2), (1,15)} & 0 & \cdots & 0 & \cdots & 0 \\
    \vdots & \vdots & & \vdots & \vdots & \vdots & & \vdots & & \vdots \\
    M_{(2,1), (1,1)} & M_{(2,1), (1,2)} & \cdots & M_{(2,1),(1,15)} & 0 & \cdots & \cdots & M_{(2,1), (2,1)} & \cdots & 0 \\
    \vdots \\
    \vdots
\end{bmatrix}
}_{ := M_{shift}}
\tilde{x}_{365}
\end{equation}
where 
$$
\tilde{x}_{365} := S^{-1} x \otimes \vb{1}_{365}
=
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
.$$

We could reorganise the vector $\tilde{x}_{365}$ to the transposed expression $\vb{1}_{365} \otimes (S^{-1} x)$, which is a vector containing all emissions on day 1, followed by all emissions on day 2, etc. A similar permutation of $y$ could be chosen. This has the effect of permuting the rows and columns of $M_{shift}$. A permutation of the variables gives a model that is essentially the same as \cref{eq: initial model 365 days} but may have advantages when it comes to implementation or computational efficiency. We will not consider this.

\subsection{Model with time-dependent correction to the emissions}
If we weigh each day and each emission source by some factor, we are applying a transformation
$$
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
\mapsto
\begin{bmatrix}
    w_{1,1} s_1^{-1} x_1 \\
    \vdots \\
    w_{1,365} s_1^{-1} x_1 \\ 
    w_{2,1} s_2^{-1} x_2 \\
    \vdots \\ 
    w_{m, 365} s_m^{-1} x_m
\end{bmatrix}
$$
which can be written concisely as 
$$
W (S^{-1} x \otimes \vb{1}_{365})
\quad\text{where}\quad 
W = \begin{bmatrix}
    w_{1,1} & 0 & \cdots & 0 \\
    0 & w_{1,2} & \cdots & 0 \\
    \vdots & \vdots & \ddots & 0 \\
    0 & 0 & \cdots & w_{m, 365}
\end{bmatrix}
.$$
Using this weighted vector of emissions instead of $\tilde{x}_{365}$ in \cref{eq: initial model 365 days} gives
\begin{equation}
    \label{eq: time-evolving model concise}
\hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
.\end{equation}
In this expression, the unknown is $W$ and should be fit to the data. 
% There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
% In the following, we provide a theoretical description of a system that can be solved for $W$.

@@

\subsection{Fitting the model}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x}_{365})}^2 + \frac{1}{2}\lambda \norm{\log W}^2
.$$
The first term measures the average logarithm of the predicted values $\hat{y}$ (detection events only). The second term measures the average logarithm of the weights.
\begin{proposition}
    Let $\hat{z}$ be the vector $[\hat{z}_1\,\,\dots\,\,\hat{z}_n]^T$ where 
    $$
    \hat{z}_i = \begin{cases}
        \frac{\log \hat{y}_i}{\hat{y}_i} & \text{if observation $i$ is a detection} \\
        0 & \text{if observation $i$ is a non-detection} 
    \end{cases}
    .$$
    Then the gradient of $f(W)$ is the diagonal matrix
    $$
    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left(M_{shift}^T \hat{z} \tilde{x}_{365}^T\right) + \lambda W^{-1} \log W
    $$
    where $\mathrm{Diag}$ is the map that sets the off-diagonal elements of a matrix to zero.
\end{proposition}

@@

Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
Define $$
G(t) :=
P_{det} \hat{y}_i
=
P_{det} M_{shift} W(t) \tilde{x}_{365}
\quad\text{and}\quad
F(t) := \log.(G(t))
.$$
Then 
$$
G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x}_{365} = (P_{det}M_{shift} \oslash \tilde{x}_{365}^T) U}_A t
$$
where $\oslash$ is the Kronecker product with reversed arguments, which satisfies $\mathrm{vec}(A B C) = (A \oslash C^T) \mathrm{vec} B$.
Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
$$
\frac{\partial}{\partial t} F(t) =
\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
.$$
Write $z := P_{det} \hat{z}$.
We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(t)$.
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
&= A^T z \\
&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}_{365}) z \\
&= U^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T \\
&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}_{365}^T \\
&= U^T M_{shift}^T z \tilde{x}_{365}^T
\end{align*}
where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
Since $U$ is unitary, we have $$
\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
UU^T (P_{det} M_{shift})^T z \tilde{x}_{365}^T
$$
The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.

>>>>

Now we present a different interpretation. Instead of $S^{-1} x \otimes \vb{1}_{15}$, we make a vector $\tilde{x} := S^{-1} x \otimes \vb{1}_{365}$. This is a vector of length $365 \times 200$ that gives the emission from day 1 to day 365 for each source. This is assumed to be constant for all days. Therefore, $\tilde{x}$ consists of $m = 200$ blocks, each of which is a constant vector of length $365$.

We can write a system of equations that is logically the same as \cref{eq: initial model concise} but uses $\tilde{x}$. This gives a larger linear system. Because the emission from most dates has a negligible contribution to the observed concentrations, many coefficients in the larger matrix will be zero. For every measurement $i$, we have a linear relationship that looks something like this:
$$
        \hat{y}_i
    =
    \begin{bmatrix}
        \cdots & M_{i,(1,1)} & \dots & M_{i,(1,15)} & 0 & \dots & M_{i,(2,1)} & \dots & M_{i,(2,15)} & \dots
    \end{bmatrix}
    (S^{-1} x \otimes \vb{1}_{365})
.
$$
The presence of zeros in the matrix indicates that emissions from the distant past do not contribute to the observed concentrations.

Now assume that in each station, there is one observation per day, named $y_{kt}$, where $k = 1,\dots,K$ where $K$ is the number of stations and $t = 1,\dots,365$. The index $t=1$ corresponds to the most recent measurement. Then the full system reads as follows:
\begin{equation}
    \label{eq: initial model 365 days}
\begin{bmatrix}
    \hat{y}_{1,1} \\
    \hat{y}_{1,2} \\
    \vdots \\
    \hat{y}_{2,1} \\ 
    \vdots \\
    \hat{y}_{K,365}
\end{bmatrix}
=
\underbrace{
\begin{bmatrix}
    M_{(1,1), (1,1)} & M_{(1,1), (1,2)} & \cdots & M_{(1,1),(1,15)} & 0 & \cdots & \cdots & M_{(1,1), (2,1)} & \cdots & 0 \\
    0 & M_{(1,2), (1,1)} & \cdots & M_{(1,2), (1,14)} & M_{(1,2), (1,15)} & 0 & \cdots & 0 & \cdots & 0 \\
    \vdots & \vdots & & \vdots & \vdots & \vdots & & \vdots & & \vdots \\
    M_{(2,1), (1,1)} & M_{(2,1), (1,2)} & \cdots & M_{(2,1),(1,15)} & 0 & \cdots & \cdots & M_{(2,1), (2,1)} & \cdots & 0 \\
    \vdots \\
    \vdots
\end{bmatrix}
}_{ := M_{shift}}
\tilde{x}
\end{equation}
where 
$$
\tilde{x} := S^{-1} x \otimes \vb{1}_{365}
=
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
.$$

We could reorganise the vector $\tilde{x}$ to the transposed expression $\vb{1}_{365} \otimes (S^{-1} x)$, which is a vector containing all emissions on day 1, followed by all emissions on day 2, etc. A similar permutation of $y$ could be chosen. This has the effect of permuting the rows and columns of $M_{shift}$. A permutation of the variables gives a model that is essentially the same as \cref{eq: initial model 365 days} but may have advantages when it comes to implementation or computational efficiency. We will not consider this.

\subsection{Model with time-dependent correction to the emissions}
If we weigh each day and each emission source by some factor, we are applying a transformation
$$
\begin{bmatrix}
    s_1^{-1} x_1 \\
    \vdots \\
    s_1^{-1} x_1 \\ 
    s_2^{-1} x_2 \\
    \vdots \\ 
    s_m^{-1} x_m
\end{bmatrix}
\mapsto
\begin{bmatrix}
    w_{1,1} s_1^{-1} x_1 \\
    \vdots \\
    w_{1,365} s_1^{-1} x_1 \\ 
    w_{2,1} s_2^{-1} x_2 \\
    \vdots \\ 
    w_{m, 365} s_m^{-1} x_m
\end{bmatrix}
$$
which can be written concisely as 
$$
W (S^{-1} x \otimes \vb{1}_{365})
\quad\text{where}\quad 
W = \begin{bmatrix}
    w_{1,1} & 0 & \cdots & 0 \\
    0 & w_{1,2} & \cdots & 0 \\
    \vdots & \vdots & \ddots & 0 \\
    0 & 0 & \cdots & w_{m, 365}
\end{bmatrix}
.$$
Using this weighted vector of emissions instead of $\tilde{x}$ in \cref{eq: initial model 365 days} gives
\begin{equation}
    \label{eq: time-evolving model concise}
\hat{y} = M_{shift} W (S^{-1} x \otimes \vb{1}_{365})
.\end{equation}
In this expression, the unknown is $W$ and should be fit to the data. 
% There are a plethora of numerical techniques for solving linear equations where the unknown is a matrix.
% In the following, we provide a theoretical description of a system that can be solved for $W$.

@@

\subsection{Fitting the model}
The indices $i = 1,\dots,n$ are partitioned as a disjoint union $\mathcal{I}_{det} \cup \mathcal{I}_{0}$ where $\mathcal{I}_0$ contains the indices of non-detection events. Let $P_{det}$ be the linear map that removes the rows of a matrix that correspond to $\mathcal{I}_0$. We can measure the accuracy as $\norm{\log.(P_{det} \hat{y})}^2$ where $\log.(\cdot)$ is the elementwise \emph{natural} logarithm. The simplicity of $W$ can be measured as $\norm{\log W}^2$. If we forget about non-detections, we have the cost function 
$$
\min_{W \text{ diagonal}} f(W) =
\underbrace{\frac{1}{2}\norm{\log.(P_{det} M_{shift} W \tilde{x})}^2}_{\text{prediction error}} + \underbrace{\frac{1}{2}\lambda \norm{\log W}^2}_{\text{deviation from average emission}}
.$$
\begin{proposition}
    Let $\hat{z}$ be the vector $[\hat{z}_1\,\,\dots\,\,\hat{z}_n]^T$ where 
    $$
    \hat{z}_i = \begin{cases}
        \frac{\log \hat{y}_i}{\hat{y}_i} & \text{if observation $i$ is a detection} \\
        0 & \text{if observation $i$ is a non-detection} 
    \end{cases}
    .$$
    Then the gradient of $f(W)$ is the diagonal matrix
    $$
    \mathrm{grad}_{W} f(W) = \mathrm{Diag}\left(M_{shift}^T \hat{z} \tilde{x}^T\right) + \lambda W^{-1} \log W
    $$
    where $\mathrm{Diag}$ is the map that sets the off-diagonal elements of a matrix to zero.
\end{proposition}

@@

Parameterise $W$ linearly as $\mathrm{vec}\, W(t) = U t$ for some unitary $U$.
Define $$
G(t) :=
P_{det} \hat{y}_i
=
P_{det} M_{shift} W(t) \tilde{x}
\quad\text{and}\quad
F(t) := \log.(G(t))
.$$
Then 
$$
G(t) = \underbrace{P_{det} M_{shift} W(t) \tilde{x} = (P_{det}M_{shift} \oslash \tilde{x}^T) U}_A t
$$
where $\oslash$ is the Kronecker product with reversed arguments, which satisfies $\mathrm{vec}(A B C) = (A \oslash C^T) \mathrm{vec} B$.
Since $G(t)$ is linear, $F(t) = \log.(G(t))$ is easy to differentiate:
$$
\frac{\partial}{\partial t} F(t) =
\frac{\partial}{\partial t} \log.(G(t)) = \mathrm{diag}\,(G(t))^{-1} A
.$$
Write $z := P_{det} \hat{z}$.
We can interpret $z$ as the vector obtained by applying the map $x \mapsto (\log x) / x$ elementwise to $G(t)$.
Using standard differentiation rules for the squared norm of a vector-valued function, we obtain
\begin{align*}
\mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right) = \left(\frac{\partial F}{\partial t}\right)^T F(t)
&= A^T \mathrm{diag}(G(t))^{-1} \log.(G(t)) \\
&= A^T z \\
&= U^T ((P_{det} M_{shift})^T \oslash \tilde{x}) z \\
&= U^T (P_{det} M_{shift})^T z \tilde{x}^T \\
&= U^T M_{shift}^T P_{det}^T P_{det} \hat{z} \tilde{x}^T \\
&= U^T M_{shift}^T z \tilde{x}^T
\end{align*}
where the last equality follows from the fact that $P_{det}^T P_{det}$ is the map that sets the rows corresponding to the non-detection events to zero. 
Since $U$ is unitary, we have $$
\mathrm{grad}_W \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
U \mathrm{grad}_t\, \left(\frac{1}{2} \norm{F(t)}^2 \right)
=
UU^T (P_{det} M_{shift})^T z \tilde{x}^T
$$
The linear map $UU^T$ is simply the orthogonal projector onto the diagonal matrices, which is equivalent to $\mathrm{Diag}$.
